SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2016-01-06 papers/B1555_EdgeOn/arXiv_0106/B1555_EdgeOn.tex
update new dir for arXiv submit

\begin{abstract}
Gravitational lens flux-ratio anomalies provide a powerful technique for measuring dark matter substructure in distant galaxies. However, before using these flux-ratio anomalies to test galaxy formation models, it is imperative to ascertain that the given anomalies are indeed due to the presence of dark matter substructure and not due to some other component of the lensing galaxy halo or to propagation effects. Here we present the case of CLASS~B1555+375, which has a strong radio-wavelength flux-ratio anomaly.  {\bf Our high-resolution near-infrared Keck~II adaptive optics imaging and archival {\it Hubble Space Telescope} data reveal the lensing galaxy in this system} to have  a clear edge-on disc component that crosses directly over the pair of images that exhibit the flux-ratio anomaly. We find simple models that include the disc can reproduce the cm-wavelength flux-ratio anomaly without requiring additional dark matter substructure. Although further studies are required, our results suggest the assumption that all flux-ratio anomalies are due to a population of dark matter sub-haloes may be incorrect, and analyses that do not account for the full complexity of the lens macro-model {\bf may} overestimate the substructure mass fraction in massive lensing galaxies.
\end{abstract}

@@

The \emph{gravitational imaging technique}, which alternatively utilizes the surface brightness distribution of extended arcs and Einstein rings to detect low-mass substructures within lensing haloes, was introduced by \citet{K05} and \citet{V09}. This method has given a measurement of the substructure mass fraction for a sample of {\bf 11 lenses} that is currently in agreement with the theoretical expectations from cold matter matter simulations \citep{V14a,V12}. The reason for the conflicting results between the flux-ratio anomaly and gravitational imaging techniques is currently not clear.  While this could be the result of small sample statistics, it could also be an indication that the flux-ratio anomalies have a different origin than clumpy substructure \citep[see][for a discussion]{Xu15}.

@@

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, 
%as can be seen by comparing the radio contours to the $K^\prime$-band emission 
(Fig.~\ref{fig:3panel}b), and the faint, but clearly visible emission from {\bf the lensing galaxy that appears to be an edge-on disc}, with a position angle of $\sim 10$~deg and an {\bf axis ratio $q \sim 0.2$}.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission from these images is likely due to dust extinction and additional demagnification due to lensing.

@@

We find that lensed images A and B have been resolved into a gravitational arc that is $\sim100$~mas in {\bf length}. Image C is still compact on mas-scales, whereas image D is only marginally detected ($\sim6~\sigma$). 
%We immediately see from the high resolution VLBA imaging 
The VLBA imaging shows that image B has a smaller angular size than image A, which is consistent with a perturbation in the mass model being responsible for the flux-ratio anomaly (recall that gravitational lensing conserves surface brightness), as opposed to free-free absorption or interstellar scattering.

@@

The next step is to test if a physically plausible edge-on disc can cause a perturbation in the strong lensing mass model that produces the observed flux ratio anomaly of  merging images A and B. An exponential disc profile best describes the disc component in spiral galaxies because it matches the light distribution \citep{Kee98}. We thus choose a SIE plus an exponential disc model (SIE+ExpDisc) in our study of B1555+375. The free parameters of the model are the SIE Einstein radius ($b$), centroid position, ellipticity ($e = 1 - q$), position angle ($\theta$), and the exponential disc intrinsic central density ($\kappa_0$) and scale length ($R_s$). A detailed definition of these mass profiles is described by \citet{Kee01}. The source position adds two additional model parameters. {\bf We have not included a third component such as an external shear or additional NFW halo term because the number of constraints provided by the compact images detected by MERLIN is not sufficient to constrain such a model.} 
%An SIE profile is also sufficient to include the halo mass since B1555 has a small Einstein radius.} 
{\bf 
More sophisticated mass models require additional
constraints, such as those from the lensed extended emission in either
the NIR or radio imaging.  While we have successfully used the former
approach in the past \citep[e.g.][]{V12}, for B1555+375 the extinction
in the AO imaging is so severe that components B and D are not
visible, and the arc that connects A and B appears to be truncated.
The low signal-to-noise ratio in the ring away from the lensed AGN and
the strong extinction in lens galaxy prevent us from using this
approach.  What are, thus, needed are significantly deeper
observations at a higher bandwidth with very long baseline
interferometric arrays, which can test more complex models in the
future by imaging the extended gravitational arc and more
significantly detecting image D. }

@@

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:3panel}c. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. The model requires that the ExpDisc component makes up about 10 per cent of the total {\bf projected} mass within the Einstein radius (1.7~kpc for an assumed lens redshift of $z_l = 1$).  {\bf In this central region of an edge-on spiral galaxy, the baryonic mass can be a significant component of the total enclosed mass \citep[e.g.][]{swells2}.  In the NIR imaging there is no sign of a bulge; a bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the ExpDisc component as a good description of the baryonic mass of the lens galaxy.}  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models require {\bf the stronger constraints provided by deep imaging of the full extent of the gravitational arc in future high resolution radio observations.}

@@

\textbf{ 
Here, we find evidence for an edge-on disc in the flux-ratio anomaly
gravitational lens B1555+375 that can account for the required
perturbation in the mass model.  Moreover, near-infrared AO imaging of
other lenses in the SHARP sample shows that B1555+375 is not the only
system that has a significant disc component detection in the lensing
galaxy; these gravitational lenses will be investigated in future
work, but the images alone suggest that other systems may have similar
links between flux-ratio anomalies and disc components. Such
high-order deviations from a smooth mass distribution have already
been suggested previously \citep{evans03, congdon05}, but this is the
first time that an observed disc has been explicitly modelled to
account for a flux-ratio anomaly. However, it is important to realise
that by including observationally-motivated modifications to a
`smooth' mass distribution, such as luminous dwarf companions or
edge-on discs, in the mass models, we are still limited in our
interpretation if we can not determine how much of the anomalies are
due to these components. For our current analysis of B1555+375, we
have shown that the entire anomaly could be caused by the edge-on
disc.  However, it may be possible to independently constrain the mass
of the disc through kinematic measurements (most likely from
mm-wavelength resolved emission line observations), which would test
whether the disc actually makes a sufficient contribution to the mass
model \citep[e.g. as in the case of B1933+503;][]{suyu12}. Also, in
the blue contours of Fig.~\ref{fig:3panel}a, B1555+375 shows evidence
for an extended $\sim$ 100 mas gravitational arc at radio wavelengths
that crosses the edge-on disc. Further high resolution imaging with
very long baseline interferometry will provide the observational
constraints required to fully test whether the disc is sufficient to
cause the anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies in some systems. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) be taken into account 
\citep[e.g.][]{Xu15}. A number of lens systems with significant disc components are
already known, including, for example, the sample of 16 edge-on disks from SDSS found in the SWELLS survey \citep{swells1}.  This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines \citep[e.g.][]{N14}.}
  %Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems \citep[and the presence of a disc-like component may not resolve the anomaly anyway;][]{evans03}, indicating that these systems are still useful for probing the dark matter substructure mass function.

>>>>

\begin{abstract}
Gravitational lens flux-ratio anomalies provide a powerful technique for measuring dark matter substructure in distant galaxies. However, before using these flux-ratio anomalies to test galaxy formation models, it is imperative to ascertain that the given anomalies are indeed due to the presence of dark matter substructure and not due to some other component of the lensing galaxy halo or to propagation effects. Here we present the case of CLASS~B1555+375, which has a strong radio-wavelength flux-ratio anomaly.  Our high-resolution near-infrared Keck~II adaptive optics imaging and archival {\it Hubble Space Telescope} data reveal the lensing galaxy in this system to have  a clear edge-on disc component that crosses directly over the pair of images that exhibit the flux-ratio anomaly. We find simple models that include the disc can reproduce the cm-wavelength flux-ratio anomaly without requiring additional dark matter substructure. Although further studies are required, our results suggest the assumption that all flux-ratio anomalies are due to a population of dark matter sub-haloes may be incorrect, and analyses that do not account for the full complexity of the lens macro-model may overestimate the substructure mass fraction in massive lensing galaxies.
\end{abstract}

@@

The \emph{gravitational imaging technique}, which alternatively utilizes the surface brightness distribution of extended arcs and Einstein rings to detect low-mass substructures within lensing haloes, was introduced by \citet{K05} and \citet{V09}. This method has given a measurement of the substructure mass fraction for a sample of 11 lenses that is currently in agreement with the theoretical expectations from cold matter matter simulations \citep{V14a,V12}. The reason for the conflicting results between the flux-ratio anomaly and gravitational imaging techniques is currently not clear.  While this could be the result of small sample statistics, it could also be an indication that the flux-ratio anomalies have a different origin than clumpy substructure \citep[see][for a discussion]{Xu15}.

@@

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, 
%as can be seen by comparing the radio contours to the $K^\prime$-band emission 
(Fig.~\ref{fig:3panel}b), and the faint, but clearly visible emission from the lensing galaxy that appears to be an edge-on disc, with a position angle of $\sim 10$~deg and an axis ratio $q \sim 0.2$.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission from these images is likely due to dust extinction and additional demagnification due to lensing.

@@

We find that lensed images A and B have been resolved into a gravitational arc that is $\sim100$~mas in length. Image C is still compact on mas-scales, whereas image D is only marginally detected ($\sim6~\sigma$). 
%We immediately see from the high resolution VLBA imaging 
The VLBA imaging shows that image B has a smaller angular size than image A, which is consistent with a perturbation in the mass model being responsible for the flux-ratio anomaly (recall that gravitational lensing conserves surface brightness), as opposed to free-free absorption or interstellar scattering.

@@

The next step is to test if a physically plausible edge-on disc can cause a perturbation in the strong lensing mass model that produces the observed flux ratio anomaly of  merging images A and B. An exponential disc profile best describes the disc component in spiral galaxies because it matches the light distribution \citep{Kee98}. We thus choose a SIE plus an exponential disc model (SIE+ExpDisc) in our study of B1555+375. The free parameters of the model are the SIE Einstein radius ($b$), centroid position, ellipticity ($e = 1 - q$), position angle ($\theta$), and the exponential disc intrinsic central density ($\kappa_0$) and scale length ($R_s$). A detailed definition of these mass profiles is described by \citet{Kee01}. The source position adds two additional model parameters. We have not included a third component such as an external shear or additional NFW halo term because the number of constraints provided by the compact images detected by MERLIN is not sufficient to constrain such a model.
%An SIE profile is also sufficient to include the halo mass since B1555 has a small Einstein radius.}

More sophisticated mass models require additional
constraints, such as those from the lensed extended emission in either
the NIR or radio imaging.  While we have successfully used the former
approach in the past \citep[e.g.][]{V12}, for B1555+375 the extinction
in the AO imaging is so severe that components B and D are not
visible, and the arc that connects A and B appears to be truncated.
The low signal-to-noise ratio in the ring away from the lensed AGN and
the strong extinction in lens galaxy prevent us from using this
approach.  What are, thus, needed are significantly deeper
observations at a higher bandwidth with very long baseline
interferometric arrays, which can test more complex models in the
future by imaging the extended gravitational arc and more
significantly detecting image D.

@@

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:3panel}c. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. The model requires that the ExpDisc component makes up about 10 per cent of the total projected mass within the Einstein radius (1.7~kpc for an assumed lens redshift of $z_l = 1$).  In this central region of an edge-on spiral galaxy, the baryonic mass can be a significant component of the total enclosed mass \citep[e.g.][]{swells2}.  In the NIR imaging there is no sign of a bulge; a bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the ExpDisc component as a good description of the baryonic mass of the lens galaxy. We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models require the stronger constraints provided by deep imaging of the full extent of the gravitational arc in future high resolution radio observations.

@@

Here, we find evidence for an edge-on disc in the flux-ratio anomaly
gravitational lens B1555+375 that can account for the required
perturbation in the mass model.  Moreover, near-infrared AO imaging of
other lenses in the SHARP sample shows that B1555+375 is not the only
system that has a significant disc component detection in the lensing
galaxy; these gravitational lenses will be investigated in future
work, but the images alone suggest that other systems may have similar
links between flux-ratio anomalies and disc components. Such
high-order deviations from a smooth mass distribution have already
been suggested previously \citep{evans03, congdon05}, but this is the
first time that an observed disc has been explicitly modelled to
account for a flux-ratio anomaly. However, it is important to realise
that by including observationally-motivated modifications to a
`smooth' mass distribution, such as luminous dwarf companions or
edge-on discs, in the mass models, we are still limited in our
interpretation if we can not determine how much of the anomalies are
due to these components. For our current analysis of B1555+375, we
have shown that the entire anomaly could be caused by the edge-on
disc.  However, it may be possible to independently constrain the mass
of the disc through kinematic measurements (most likely from
mm-wavelength resolved emission line observations), which would test
whether the disc actually makes a sufficient contribution to the mass
model \citep[e.g. as in the case of B1933+503;][]{suyu12}. Also, in
the blue contours of Fig.~\ref{fig:3panel}a, B1555+375 shows evidence
for an extended $\sim$ 100 mas gravitational arc at radio wavelengths
that crosses the edge-on disc. Further high resolution imaging with
very long baseline interferometry will provide the observational
constraints required to fully test whether the disc is sufficient to
cause the anomaly.

Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies in some systems. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) be taken into account 
\citep[e.g.][]{Xu15}. A number of lens systems with significant disc components are
already known, including, for example, the sample of 16 edge-on disks from SDSS found in the SWELLS survey \citep{swells1}.  This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines \citep[e.g.][]{N14}.
  %Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems \citep[and the presence of a disc-like component may not resolve the anomaly anyway;][]{evans03}, indicating that these systems are still useful for probing the dark matter substructure mass function.
